---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(balancing.quiet = TRUE)
```

# balancing

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-causal/balancing/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-causal/balancing/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/r-causal/balancing/graph/badge.svg)](https://app.codecov.io/gh/r-causal/balancing)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

balancing calculates optimization-based balancing weights for causal inference. Rather than modeling the probability of exposure and then hoping the resulting weights balance the covariates, these methods make covariate balance the objective of a convex optimization problem and solve for the weights that achieve it directly. The package covers six methods (entropy balancing, inverse probability tilting, the covariate balancing propensity score, energy balancing, characteristic function distance balancing, and stable balancing weights) across binary, categorical, and continuous exposures and a range of estimands. A Rust core provides the numerical solvers.

## Installation

You can install the development version of balancing from [GitHub](https://github.com/r-causal/balancing) with:

``` r
# install.packages("pak")
pak::pak("r-causal/balancing")
```

Installing from source requires a Rust toolchain (`rustc` 1.85 or newer and Cargo).

## Usage

`balance()` is the entry point. You give it a data frame, name the exposure and covariates with tidyselect, choose a method and an estimand, and it returns a fitted object carrying the weights and a balance table.

```{r example}
library(balancing)

# Simulate data with two confounders and a binary exposure
set.seed(1)
n <- 500
x1 <- rnorm(n)
x2 <- rnorm(n)
z <- rbinom(n, 1, plogis(0.4 * x1 - 0.5 * x2))
y <- 1 + 0.9 * z + 0.6 * x1 - 0.4 * x2 + rnorm(n)
study <- data.frame(exposure = z, age = x1, score = x2, outcome = y)

# Fit entropy balancing weights for the ATT
fit <- balance(
  study,
  exposure,
  c(age, score),
  method = bw_entropy(),
  estimand = "att"
)

fit
```

The printed summary reports the solver status and the largest imbalance the weights leave behind on the constraint terms.

The weights are a `bw` vector, a sibling of `propensity::psw`. Pass them to a weighted outcome model to estimate the effect.

```{r outcome}
study$w <- as.numeric(weights(fit))
outcome_mod <- lm(outcome ~ exposure, data = study, weights = w)
coef(outcome_mod)[["exposure"]]
```

For the estimating-equation methods (entropy balancing, inverse probability tilting, and the covariate balancing propensity score), pass the fit and the weighted outcome model to `ipw()` to get effect estimates with standard errors that account for having estimated the weights. Binary, categorical, and continuous exposures are supported; with a continuous exposure the reported effect is the exposure coefficient of a weighted marginal structural model.

```{r ipw}
ipw(fit, outcome_mod)
```

## How balancing relates to the other r-causal packages

balancing is part of the [r-causal](https://github.com/r-causal) family and shares its design language with three sibling packages.

- [propensity](https://r-causal.github.io/propensity/) calculates inverse probability weights from a fitted propensity score model and estimates effects with `ipw()`. balancing produces a different kind of weight, one that targets balance directly rather than through a modeled score, but it plugs into the same effect-estimation workflow: the weights share the `causal_wts` vocabulary, and `ipw()` dispatches on a balancing fit just as it does on a propensity model.
- [halfmoon](https://r-causal.github.io/halfmoon/) assesses covariate balance. balancing reports the balance achieved on its own constraint terms, but a full balance-assessment workflow, including variables the weights were not asked to balance, lives in halfmoon. Extract the weights with `weights()` and pass them to `halfmoon::check_balance()` and `halfmoon::plot_balance()`.
- [positively](https://r-causal.github.io/positively/) diagnoses positivity violations and extrapolation across binary, categorical, and continuous exposures. balancing redistributes influence across units to achieve balance, and positivity problems are exactly where that redistribution becomes extrapolation, showing up as extreme weights and a collapsing effective sample size, so checking positivity with positively complements weighting with balancing.

## Learn more

- `vignette("balancing")` walks through a binary exposure workflow from weights to effects.
- `vignette("choosing-a-method")` compares the six methods and the constraint options.
- `vignette("inference")` covers standard errors after balancing.
- [Causal Inference in R](https://www.r-causal.org/) is a book on causal inference methods in R.
