---
title: "Inference after balancing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference after balancing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(balancing.quiet = TRUE)
set.seed(2024)
```

Balancing weights are estimated, not fixed. A weighted outcome model treats its weights as known constants, so the standard errors it reports do not account for the uncertainty in the weights themselves. This article explains why that matters and shows two ways to get honest standard errors: M-estimation through `ipw()` for the estimating-equation methods, and a bootstrap for everything else.

```{r setup}
library(balancing)
```

## Why the weights' uncertainty matters

Consider a binary exposure with a binary outcome and two confounders.

```{r data}
n <- 1500
age <- rnorm(n)
score <- rnorm(n)
exposure <- rbinom(n, 1, plogis(0.5 * age - 0.5 * score))
event <- rbinom(n, 1, plogis(-0.4 + 0.7 * exposure + 0.4 * age - 0.3 * score))

study <- data.frame(exposure, age, score, event)
```

We fit entropy balancing weights for the ATE and a weighted outcome model.

```{r fit}
fit <- balance(
  study,
  exposure,
  c(age, score),
  method = bw_entropy(),
  estimand = "ate"
)
study$w <- weights(fit)

outcome_mod <- glm(
  event ~ exposure,
  data = study,
  family = quasibinomial(),
  weights = w
)
```

The outcome model reports a standard error for the exposure coefficient, but it is conditional on the weights. It treats the reweighting as given and so leaves out one of the sources of variation in the estimate.

```{r naive}
naive <- summary(outcome_mod)$coefficients["exposure", ]
naive
```

The right standard error propagates the uncertainty from estimating the weights into the effect estimate. Whether it is larger or smaller than the naive value depends on the estimand and the design, but it is generally not the same, and using the naive value can give intervals with the wrong coverage.

## M-estimation with `ipw()`

The estimating-equation methods (`bw_entropy()`, `bw_ipt()`, and the just-identified `bw_cbps()`) fit weights that solve smooth estimating equations. That structure lets the uncertainty be propagated analytically, without resampling. `ipw()` writes one stacked estimating-equation system whose parameters are the weight parameters, the outcome-model coefficients, the two marginal means, and the effect contrasts, and the deli package differentiates that system at the fitted values and returns its empirical sandwich covariance. Nothing is re-solved: every parameter enters at the value its own fit already found. Because each contrast is a parameter of the stack rather than a transformation applied afterward, its standard error comes straight off the diagonal of the joint covariance, with no delta-method step in between. The result is a standard error that accounts for having estimated the weights.

```{r ipw}
result <- ipw(fit, outcome_mod)
result
```

The `rd` row is the risk difference on the same scale as the outcome model's marginal effect. Comparing its standard error to the naive one shows the correction.

```{r compare}
rd <- as.data.frame(result$estimates)
rd <- rd[rd$effect == "rd", ]

data.frame(
  source = c("naive (weights fixed)", "ipw (weights estimated)"),
  std.err = c(naive[["Std. Error"]], rd$std.err)
)
```

`ipw()` computes this variance for an estimating-equation fit at exact balance: any of the three methods at a binary or categorical exposure, and entropy balancing at a continuous one. The outcome model may be a `glm()` with a binomial, quasibinomial, or gaussian family, or a plain `lm()`. For a discrete exposure any link those families carry is handled exactly, including a non-canonical one such as probit, because the sandwich differentiates the score it is given rather than assuming an information matrix. Model offsets are supported natively, written either as an `offset()` term in the formula or passed through the model's `offset` argument, and are carried through both the outcome-model score and the marginal means. A continuous exposure reports a coefficient rather than a contrast of predictions, so an offset that reads the exposure moves that coefficient off the effect it names and is refused there; an exposure-free offset, such as the person-time offset of a rate model, is supported as it is everywhere else. That refusal reads the offset expression, so an offset arriving as a precomputed vector is beyond it and keeping the exposure out of one is yours to honor.

### Categorical exposures

A categorical exposure works the same way, with one marginal mean per level. The stacked system carries all of them, and the reported effects are the contrasts of each non-reference level against the reference level, which is the first of the exposure's own levels. The outcome model enters the exposure as a factor.

```{r categorical}
odds_medium <- exp(0.6 * age - 0.3 * score)
odds_high <- exp(-0.4 * age + 0.7 * score)
denominator <- 1 + odds_medium + odds_high
draw <- runif(n)

study$arm <- factor(
  ifelse(
    draw < 1 / denominator,
    "low",
    ifelse(draw < (1 + odds_medium) / denominator, "medium", "high")
  ),
  levels = c("low", "medium", "high")
)
study$relapse <- rbinom(
  n,
  1,
  plogis(
    -0.4 +
      0.5 * (study$arm == "medium") +
      0.9 * (study$arm == "high") +
      0.4 * age -
      0.3 * score
  )
)

arm_fit <- balance(
  study,
  arm,
  c(age, score),
  method = bw_ipt(),
  estimand = "ate"
)
study$arm_w <- weights(arm_fit)

arm_mod <- glm(
  relapse ~ arm,
  data = study,
  family = quasibinomial(),
  weights = arm_w
)

ipw(arm_fit, arm_mod)
```

Each effect measure appears once per comparison, so the estimates table gains a `comparison` column naming the two levels each row contrasts. Everything else carries over unchanged: the outcome model may adjust for covariates, the marginal means standardize over the estimand's target population, and the standard errors account for having estimated the weights.

### Continuous exposures

A continuous exposure has no levels to contrast, so there is no pair of marginal means to difference. `ipw()` reports the dose-response coefficient of a weighted marginal structural model instead: entropy balancing removes the association between the exposure and the covariates, the outcome model carries exactly one term in the exposure, and that term's coefficient is the effect of a one-unit change in the exposure on the model's own link scale.

```{r continuous}
study$dose <- 0.8 * age - 0.5 * score + rnorm(n)
study$response <- 1 + 0.4 * study$dose + 0.5 * age - 0.3 * score + rnorm(n)

dose_fit <- balance(
  study,
  dose,
  c(age, score),
  method = bw_entropy(),
  estimand = "ate"
)
study$dose_w <- weights(dose_fit)

dose_mod <- lm(response ~ dose, data = study, weights = dose_w)

ipw(dose_fit, dose_mod)
```

The table holds one row, named for the link: `slope` for an identity link, `log(or)` for a logit, and `log(rr)` for a log link. The model may adjust for covariates, but exactly one of its terms may read the exposure: `response ~ dose + age` is accepted, while `response ~ poly(dose, 2)` and `response ~ dose * age` are refused, since no single coefficient of either is the effect the row would name. Only entropy balancing at exact balance reaches this path, and only for the ATE, which is the only estimand a continuous fit targets.

The continuous standard error reaches its nominal coverage more slowly than the binary risk difference does. Over 500 draws, its ratio of mean standard error to the standard deviation of the estimates was 0.836 at 300 observations and 0.942 at 1200, so an interval at a few hundred observations is too narrow. Bootstrap it at that size.

### Covariate-adjusted outcome models

The outcome model must carry the exposure among its predictors, but it may adjust for covariates alongside it, and may interact them with the exposure. Adjusting for the same covariates the weights balance is a common way to reduce residual confounding and tighten the estimate.

```{r adjusted}
adjusted_mod <- glm(
  event ~ exposure + age + score,
  data = study,
  family = quasibinomial(),
  weights = w
)

ipw(fit, adjusted_mod)
```

Adjustment changes what the marginal means average over. A marginal model predicts one value per exposure level, so it makes no difference which units you average across; an adjusted model predicts a value per unit, and the population you average across becomes part of the estimand. `ipw()` standardizes over the estimand's target population: every unit for the ATE, and the focal group's units for the ATT or the ATC. Sampling weights, where the fit has them, weight that average as well. The standard error accounts for the adjustment the same way it accounts for everything else, by carrying the adjusted model's own score equations in the stack.

A model that leaves the exposure out has nothing to contrast, so it is refused rather than reported as a null effect.

```{r no-exposure, error = TRUE}
ipw(fit, glm(event ~ age + score, data = study, family = quasibinomial(), weights = w))
```

Everything else requires the bootstrap. The quadratic-program methods (`bw_energy()`, `bw_cfd()`, and `bw_sbw()`) and an entropy fit at a positive tolerance solve inequality-constrained problems, whose solutions are characterized by which constraints bind at the optimum rather than by a smooth system set to zero. An over-identified `bw_cbps()` fit minimizes a generalized-method-of-moments criterion in more moment conditions than it has parameters, so its balancing conditions are not solved to zero either. None of these has an estimating-equation representation to stack, so the fit carries none and `ipw()` raises an error pointing here. Inverse probability tilting is the exception to the tolerance rule: it solves the same smooth estimating equations whatever tolerance is requested, so `ipw()` handles it at any tolerance.

```{r unsupported, error = TRUE}
sbw_fit <- balance(
  study,
  exposure,
  c(age, score),
  method = bw_sbw(),
  constraints = balance_terms(tolerance = 0.02)
)
ipw(sbw_fit, outcome_mod)
```

## Bootstrapping the other fits

Where a fit carries no estimating equations, inference uses resampling instead. The recipe is a standard nonparametric bootstrap: resample the rows with replacement, refit the weights and the outcome model on each resample, and take the spread of the effect estimates across resamples as the standard error. The important detail is that the reweighting is refit inside the loop, so every source of variation, including the estimation of the weights, is captured. The example below uses `bw_sbw()`, but the recipe is the same for any fit and it also covers the cases `ipw()` declines for reasons other than the weights, such as an outcome model outside the supported families.

We'll show a hand-rolled approach, but there are several tools for bootstrapping in R. See in particular the `{boot}` package and the `{rsample}` package. See also the appendix on the bootstrap in [Causal Inference in R](https://www.r-causal.org/appendices/a-bootstrap).

We keep the example small so it runs quickly: a few hundred resamples on a modest sample.

```{r bootstrap}
boot_study <- study[seq_len(400), ]

fit_effect <- function(data) {
  fit <- balance(
    data,
    exposure,
    c(age, score),
    method = bw_sbw(),
    constraints = balance_terms(tolerance = 0.02)
  )
  data$w <- weights(fit)
  model <- glm(
    event ~ exposure,
    data = data,
    family = quasibinomial(),
    weights = w
  )
  # The marginal risk difference by g-computation
  p1 <- predict(model, transform(data, exposure = 1), type = "response")
  p0 <- predict(model, transform(data, exposure = 0), type = "response")
  mean(p1) - mean(p0)
}

set.seed(1)
point_estimate <- fit_effect(boot_study)

n_boot <- 200
boot_estimates <- vapply(seq_len(n_boot), function(b) {
  rows <- sample(nrow(boot_study), replace = TRUE)
  fit_effect(boot_study[rows, ])
}, numeric(1))
```

The bootstrap standard error is the standard deviation of the resampled estimates, and a percentile interval comes from their quantiles.

```{r bootstrap-summary}
data.frame(
  estimate = point_estimate,
  std.err = sd(boot_estimates),
  ci.lower = quantile(boot_estimates, 0.025),
  ci.upper = quantile(boot_estimates, 0.975),
  row.names = NULL
)
```

A real analysis would use more resamples, typically at least one or two thousand, and might use the bias-corrected and accelerated interval rather than the percentile one. The structure stays the same: refit the weights inside every resample.

## Choosing an approach

Use `ipw()` when the method carries estimating equations, which is the estimating-equation family with a binary or categorical exposure and exact balance, and entropy balancing with exact balance at a continuous exposure. It is faster and needs no tuning. Use the bootstrap for the quadratic-program methods, for an entropy fit at a positive tolerance, for an over-identified `bw_cbps()` fit, and for a continuous fit from a method that solves no estimating equations. A continuous exposure is worth bootstrapping at a small sample size in any case: its stacked standard error is a large-sample one and runs anticonservative at a few hundred observations. Covariate-adjusted outcome models need neither route in particular: `ipw()` stacks them directly. The bootstrap is more general but more expensive, and its precision improves with the number of resamples.
