---
title: "Inference after balancing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference after balancing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(balancing.quiet = TRUE)
set.seed(2024)
```

Balancing weights are estimated, not fixed. A weighted outcome model treats its weights as known constants, so the standard errors it reports do not account for the uncertainty in the weights themselves. This article explains why that matters and shows two ways to get honest standard errors: M-estimation through `ipw()` for the estimating-equation methods, and a bootstrap for everything else.

```{r setup}
library(balancing)
```

## Why the weights' uncertainty matters

Consider a binary exposure with a binary outcome and two confounders.

```{r data}
n <- 1500
age <- rnorm(n)
score <- rnorm(n)
exposure <- rbinom(n, 1, plogis(0.5 * age - 0.5 * score))
event <- rbinom(n, 1, plogis(-0.4 + 0.7 * exposure + 0.4 * age - 0.3 * score))

study <- data.frame(exposure, age, score, event)
```

We fit entropy balancing weights for the ATE and a weighted outcome model.

```{r fit}
fit <- balance(
  study,
  exposure,
  c(age, score),
  method = bw_entropy(),
  estimand = "ate"
)
study$w <- weights(fit)

outcome_mod <- glm(
  event ~ exposure,
  data = study,
  family = quasibinomial(),
  weights = w
)
```

The outcome model reports a standard error for the exposure coefficient, but it is conditional on the weights. It treats the reweighting as given and so leaves out one of the sources of variation in the estimate.

```{r naive}
naive <- summary(outcome_mod)$coefficients["exposure", ]
naive
```

The right standard error propagates the uncertainty from estimating the weights into the effect estimate. Whether it is larger or smaller than the naive value depends on the estimand and the design, but it is generally not the same, and using the naive value can give intervals with the wrong coverage.

## M-estimation with `ipw()`

The estimating-equation methods (`bw_entropy()`, `bw_ipt()`, and the just-identified `bw_cbps()`) fit weights that solve smooth estimating equations. That structure lets the uncertainty be propagated analytically, without resampling. `ipw()` writes one stacked estimating-equation system whose parameters are the weight parameters, the outcome-model coefficients, the two marginal means, and the effect contrasts, and the deli package differentiates that system at the fitted values and returns its empirical sandwich covariance. Nothing is re-solved: every parameter enters at the value its own fit already found. Because each contrast is a parameter of the stack rather than a transformation applied afterward, its standard error comes straight off the diagonal of the joint covariance, with no delta-method step in between. The result is a standard error that accounts for having estimated the weights.

```{r ipw}
result <- ipw(fit, outcome_mod)
result
```

The `rd` row is the risk difference on the same scale as the outcome model's marginal effect. Comparing its standard error to the naive one shows the correction.

```{r compare}
rd <- as.data.frame(result$estimates)
rd <- rd[rd$effect == "rd", ]

data.frame(
  source = c("naive (weights fixed)", "ipw (weights estimated)"),
  std.err = c(naive[["Std. Error"]], rd$std.err)
)
```

`ipw()` computes this variance for an estimating-equation fit at exact balance: any of the three methods at a binary or categorical exposure, and entropy balancing at a continuous one. The outcome model may be a `glm()` with a binomial, quasibinomial, or gaussian family, or a plain `lm()`. For a discrete exposure any link those families carry is handled exactly, including a non-canonical one such as probit, because the sandwich differentiates the score it is given rather than assuming an information matrix. Model offsets are supported natively, written either as an `offset()` term in the formula or passed through the model's `offset` argument, and are carried through both the outcome-model score and the marginal means. A continuous exposure reports coefficients rather than contrasts of predictions, so an offset that reads the exposure moves them off the effects they name and is refused there; an exposure-free offset, such as the person-time offset of a rate model, is supported as it is everywhere else. That refusal reads the offset expression, so an offset arriving as a precomputed vector is beyond it and keeping the exposure out of one is yours to honor.

### Categorical exposures

A categorical exposure works the same way, with one marginal mean per level. The stacked system carries all of them, and the table reports each of those means and then, for a plain categorical exposure, the contrasts of each non-reference level against the reference level, which is the first of the exposure's own levels. A crossing of two treatments declared with `causalgenerics::joint_exposure()` is reported in the treatments rather than in the cells, and the section on joint exposures below describes those rows. The outcome model enters the exposure as a factor.

```{r categorical}
odds_medium <- exp(0.6 * age - 0.3 * score)
odds_high <- exp(-0.4 * age + 0.7 * score)
denominator <- 1 + odds_medium + odds_high
draw <- runif(n)

study$arm <- factor(
  ifelse(
    draw < 1 / denominator,
    "low",
    ifelse(draw < (1 + odds_medium) / denominator, "medium", "high")
  ),
  levels = c("low", "medium", "high")
)
study$relapse <- rbinom(
  n,
  1,
  plogis(
    -0.4 +
      0.5 * (study$arm == "medium") +
      0.9 * (study$arm == "high") +
      0.4 * age -
      0.3 * score
  )
)

arm_fit <- balance(
  study,
  arm,
  c(age, score),
  method = bw_ipt(),
  estimand = "ate"
)
study$arm_w <- weights(arm_fit)

arm_mod <- glm(
  relapse ~ arm,
  data = study,
  family = quasibinomial(),
  weights = arm_w
)

ipw(arm_fit, arm_mod)
```

The table leads with the marginal mean of each arm and then reports each effect measure once per contrast, and the `contrast` column names what each row belongs to: the level itself for a mean, and the two levels compared for an effect. A binary exposure is read by the same rule, its means named for the two levels and its single comparison named as a contrast of them. Everything else carries over unchanged: the outcome model may adjust for covariates, the marginal means standardize over the estimand's target population, and the standard errors account for having estimated the weights.

### Effect modification with `.by`

`.by` names a modifier, and the result reports the rows it would report without one, then the marginal means and effects within each level of the modifier, then each non-reference subgroup against the reference one. The table gains a `group` column naming the subgroup a row was estimated in, placed after `contrast`, since a subgroup qualifies a whole comparison rather than one side of it.

The data below carry a treatment whose effect is larger among men than among women by construction.

```{r by-data}
set.seed(7)
m <- 500
severity <- rnorm(m)
comorbidity <- rnorm(m)
sex <- factor(
  sample(c("female", "male"), m, replace = TRUE),
  levels = c("female", "male")
)
treated <- rbinom(m, 1, plogis(0.5 * severity - 0.4 * comorbidity))
recovery <- rbinom(
  m,
  1,
  plogis(
    -0.4 +
      0.45 * treated +
      0.9 * treated * (sex == "male") +
      0.4 * severity -
      0.3 * comorbidity
  )
)

trial <- data.frame(treated, severity, comorbidity, sex, recovery)
```

The weights are fit once, over the whole sample, and the outcome model interacts the exposure with the modifier so that the fitted response is free to differ between the two groups.

```{r by}
trial_fit <- balance(
  trial,
  treated,
  c(severity, comorbidity),
  method = bw_entropy(),
  estimand = "ate"
)
trial$trial_w <- weights(trial_fit)

trial_mod <- glm(
  recovery ~ treated * sex,
  data = trial,
  family = quasibinomial(),
  weights = trial_w
)

ipw(trial_fit, trial_mod, .by = sex)
```

The `overall` block is the result an ungrouped call reports, and the blocks after it are the effect within each level of `sex`. A subgroup's marginal means are the g-computation means over that subgroup's units alone, and its risk difference is the contrast of those two means. Nothing is refit inside a subgroup: the weights are the ones the single fit produced, and no subgroup gets balance conditions of its own. The last two rows compare the subgroups, and since every row here is a parameter of the same stacked system, that comparison is a parameter as well rather than a difference taken between two separately fitted effects. Its standard error therefore carries the covariance between the two subgroups, which share the weight parameters and the outcome model's coefficients; adding two independently estimated variances would miss that covariance, in a direction the design decides.

The log odds ratio stays among the whole-sample rows. It is noncollapsible, so the odds ratio over a sample is not an average of the odds ratios within its subgroups, and a difference of two of them moves with each subgroup's baseline risk whether or not the effect is modified at all. The collapsible measures, `rd` and `log(rr)`, are the ones reported for the subgroups and for their contrast.

The subgroup contrast is not what the interaction term in `recovery ~ treated * sex` reports. That coefficient belongs to the outcome model's conditional surface: it is the change in the model's log odds ratio between the two levels of `sex`, on the link scale the model was fitted with. The `.by` rows are an estimand instead, the marginal effect of treating standardized over each subgroup's own covariate distribution, and their contrast is the difference between two such effects on each collapsible scale. The two are different quantities and generally different numbers, so a term in the model is not a reading of the request and the request is not a reading of the term. What `.by` does need from the model is room for the effect to differ. The subgroup rows are g-computation on the model as it was specified, so a model carrying no term that reads both the exposure and the modifier is warned about rather than refused: the modification may reach it through a column derived from the modifier instead.

### Joint exposures

Intervening on two treatments at once is a third question again, and it is not the one `.by` answers. `causalgenerics::joint_exposure()` crosses two discrete treatments into one categorical exposure and records the crossing on the column. `balance()` weights that column as it weights any factor over the same cells, so what the declaration changes is which effects are reported rather than which population is balanced.

```{r joint-data}
set.seed(21)
k <- 500
frailty <- rnorm(k)
drug <- rbinom(k, 1, plogis(0.4 * frailty))
diet <- rbinom(k, 1, plogis(-0.2 + 0.5 * frailty + 0.6 * drug))
remission <- rbinom(
  k,
  1,
  plogis(-1.4 + 0.3 * drug + 0.3 * diet + 1.6 * drug * diet + 0.5 * frailty)
)

regimens <- data.frame(
  frailty = frailty,
  remission = remission,
  drug = factor(drug),
  diet = factor(diet)
)
# Built after the frame so the crossing reads the same two columns the frame
# carries.
regimens$regimen <- causalgenerics::joint_exposure(
  drug = regimens$drug,
  diet = regimens$diet
)
```

```{r joint}
regimen_fit <- balance(
  regimens,
  regimen,
  c(frailty),
  method = bw_ipt(),
  estimand = "ate"
)
regimens$regimen_w <- weights(regimen_fit)

regimen_mod <- glm(
  remission ~ regimen + frailty,
  data = regimens,
  family = quasibinomial(),
  weights = regimen_w
)

ipw(regimen_fit, regimen_mod)
```

The table leads with the counterfactual mean of each cell, one per combination of the two treatments, each standardized over the whole sample. The simple effects follow: each treatment against its own reference level, within a level of the other that the `group` column names. The last pair of rows is the interaction, the difference between the drug's effect at `diet = 1` and its effect at `diet = 0`, which is large here because each treatment does much less on its own than the two do together. Reported as a plain categorical exposure, these four cells would give each cell against the reference cell, under contrasts like `"drug = 1, diet = 0 vs drug = 0, diet = 0"`; the declaration replaces those rows rather than adding to them, so no row here is named that way. The interaction appears once, under the first treatment's framing, because it is symmetric in the two: the change in the drug's effect across the levels of the diet is the change in the diet's effect across the levels of the drug, and reporting it twice would put one quantity in the table under two names. No contrast row carries a log odds ratio, for the reason no subgroup row does.

The declaration is read off the exposure column of the frame `ipw()` resolves, which is the outcome model's own frame here, so dropping it with `factor()` returns the cell-against-cell rows. A declared crossing is reported for the ATE alone, and `.by` on one is refused: effect modification of a joint intervention is a three-way question, and this surface reports neither it nor a projection of it. `?ipw.balancing` gives the full row set and both restrictions.

### Continuous exposures

A continuous exposure has no levels to contrast, so there is no pair of marginal means to difference. `ipw()` reports the dose response of a weighted marginal structural model instead: entropy balancing removes the association between the exposure and the covariates, and every coefficient of the outcome model that reads the exposure is an effect on the model's own link scale.

```{r continuous}
study$dose <- 0.8 * age - 0.5 * score + rnorm(n)
study$response <- 1 + 0.4 * study$dose + 0.5 * age - 0.3 * score + rnorm(n)

dose_fit <- balance(
  study,
  dose,
  c(age, score),
  method = bw_entropy(),
  estimand = "ate"
)
study$dose_w <- weights(dose_fit)

dose_mod <- lm(response ~ dose, data = study, weights = dose_w)

ipw(dose_fit, dose_mod)
```

An exposure entering through one design column is the whole of the dose response, so its coefficient is that response's slope everywhere. The table holds one row, named for the link: `slope` for an identity link, `log(or)` for a logit, and `log(rr)` for a log link.

The exposure may also enter through several columns. What the model has to keep is variable membership: every term reading the exposure must read the exposure alone, however many columns it expands to, so a curve written out term by term and one handed to a basis constructor are both reported.

```{r continuous-basis}
curve_mod <- lm(response ~ splines::ns(dose, 3), data = study, weights = dose_w)

ipw(dose_fit, curve_mod)
```

The table then holds one row per coefficient and gains a `contrast` column naming each row after the coefficient the fit names. The scale word steps back to `coef` at an identity link, since a curve has a different slope at every dose and no one of its coefficients is that slope; a logit still reports `log(or)` and a log link `log(rr)`, because a coefficient of those models is a log ratio whatever column it multiplies. Nothing is standardized on this path either way, so each row is a coefficient of the weighted fit and what the stack adds is the standard error.

A term reading a covariate alongside the exposure is refused. `response ~ dose * age` contributes a coefficient that is a change in the dose response per unit of `age`, so there is no one effect for a row to report and no value of `age` a row could name it at. Only entropy balancing at exact balance reaches this path, and only for the ATE, which is the only estimand a continuous fit targets.

The continuous standard error reaches its nominal coverage more slowly than the binary risk difference does. Over 500 draws, its ratio of mean standard error to the standard deviation of the estimates was 0.836 at 300 observations and 0.942 at 1200, so an interval at a few hundred observations is too narrow. Bootstrap it at that size.

### Covariate-adjusted outcome models

The outcome model must carry the exposure among its predictors, but it may adjust for covariates alongside it, and may interact them with the exposure. Adjusting for the same covariates the weights balance is a common way to reduce residual confounding and tighten the estimate.

```{r adjusted}
adjusted_mod <- glm(
  event ~ exposure + age + score,
  data = study,
  family = quasibinomial(),
  weights = w
)

ipw(fit, adjusted_mod)
```

Adjustment changes what the marginal means average over. A marginal model predicts one value per exposure level, so it makes no difference which units you average across; an adjusted model predicts a value per unit, and the population you average across becomes part of the estimand. `ipw()` standardizes over the estimand's target population: every unit for the ATE, and the focal group's units for the ATT or the ATC. Sampling weights, where the fit has them, weight that average as well. The standard error accounts for the adjustment the same way it accounts for everything else, by carrying the adjusted model's own score equations in the stack.

A model that leaves the exposure out has nothing to contrast, so it is refused rather than reported as a null effect.

```{r no-exposure, error = TRUE}
ipw(fit, glm(event ~ age + score, data = study, family = quasibinomial(), weights = w))
```

Everything else requires the bootstrap. The quadratic-program methods (`bw_energy()`, `bw_cfd()`, and `bw_sbw()`) and an entropy fit at a positive tolerance solve inequality-constrained problems, whose solutions are characterized by which constraints bind at the optimum rather than by a smooth system set to zero. An over-identified `bw_cbps()` fit minimizes a generalized-method-of-moments criterion in more moment conditions than it has parameters, so its balancing conditions are not solved to zero either. None of these has an estimating-equation representation to stack, so the fit carries none and `ipw()` raises an error pointing here. Inverse probability tilting is the exception to the tolerance rule: it solves the same smooth estimating equations whatever tolerance is requested, so `ipw()` handles it at any tolerance.

```{r unsupported, error = TRUE}
sbw_fit <- balance(
  study,
  exposure,
  c(age, score),
  method = bw_sbw(),
  constraints = balance_terms(tolerance = 0.02)
)
ipw(sbw_fit, outcome_mod)
```

## Bootstrapping the other fits

Where a fit carries no estimating equations, inference uses resampling instead. The recipe is a standard nonparametric bootstrap: resample the rows with replacement, refit the weights and the outcome model on each resample, and take the spread of the effect estimates across resamples as the standard error. The important detail is that the reweighting is refit inside the loop, so every source of variation, including the estimation of the weights, is captured. The example below uses `bw_sbw()`, but the recipe is the same for any fit and it also covers the cases `ipw()` declines for reasons other than the weights, such as an outcome model outside the supported families.

We'll show a hand-rolled approach, but there are several tools for bootstrapping in R. See in particular the `{boot}` package and the `{rsample}` package. See also the appendix on the bootstrap in [Causal Inference in R](https://www.r-causal.org/appendices/a-bootstrap).

We keep the example small so it runs quickly: a few hundred resamples on a modest sample.

```{r bootstrap}
boot_study <- study[seq_len(400), ]

fit_effect <- function(data) {
  fit <- balance(
    data,
    exposure,
    c(age, score),
    method = bw_sbw(),
    constraints = balance_terms(tolerance = 0.02)
  )
  data$w <- weights(fit)
  model <- glm(
    event ~ exposure,
    data = data,
    family = quasibinomial(),
    weights = w
  )
  # The marginal risk difference by g-computation
  p1 <- predict(model, transform(data, exposure = 1), type = "response")
  p0 <- predict(model, transform(data, exposure = 0), type = "response")
  mean(p1) - mean(p0)
}

set.seed(1)
point_estimate <- fit_effect(boot_study)

n_boot <- 200
boot_estimates <- vapply(seq_len(n_boot), function(b) {
  rows <- sample(nrow(boot_study), replace = TRUE)
  fit_effect(boot_study[rows, ])
}, numeric(1))
```

The bootstrap standard error is the standard deviation of the resampled estimates, and a percentile interval comes from their quantiles.

```{r bootstrap-summary}
data.frame(
  estimate = point_estimate,
  std.err = sd(boot_estimates),
  ci.lower = quantile(boot_estimates, 0.025),
  ci.upper = quantile(boot_estimates, 0.975),
  row.names = NULL
)
```

A real analysis would use more resamples, typically at least one or two thousand, and might use the bias-corrected and accelerated interval rather than the percentile one. The structure stays the same: refit the weights inside every resample.

## Choosing an approach

Use `ipw()` when the method carries estimating equations, which is the estimating-equation family with a binary or categorical exposure and exact balance, and entropy balancing with exact balance at a continuous exposure. It is faster and needs no tuning. Use the bootstrap for the quadratic-program methods, for an entropy fit at a positive tolerance, for an over-identified `bw_cbps()` fit, and for a continuous fit from a method that solves no estimating equations. A continuous exposure is worth bootstrapping at a small sample size in any case: its stacked standard error is a large-sample one and runs anticonservative at a few hundred observations. Covariate-adjusted outcome models need neither route in particular: `ipw()` stacks them directly. The bootstrap is more general but more expensive, and its precision improves with the number of resamples.
