---
title: "Inference after balancing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference after balancing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(balancing.quiet = TRUE)
set.seed(2024)
```

Balancing weights are estimated, not fixed. A weighted outcome model treats its weights as known constants, so the standard errors it reports do not account for the uncertainty in the weights themselves. This article explains why that matters and shows two ways to get honest standard errors: M-estimation through `ipw()` for the estimating-equation methods, and a bootstrap for everything else.

```{r setup}
library(balancing)
```

## Why the weights' uncertainty matters

Consider a binary exposure with a binary outcome and two confounders.

```{r data}
n <- 1500
age <- rnorm(n)
score <- rnorm(n)
exposure <- rbinom(n, 1, plogis(0.5 * age - 0.5 * score))
event <- rbinom(n, 1, plogis(-0.4 + 0.7 * exposure + 0.4 * age - 0.3 * score))

study <- data.frame(exposure, age, score, event)
```

We fit entropy balancing weights for the ATE and a weighted outcome model.

```{r fit}
fit <- balance(
  study,
  exposure,
  c(age, score),
  method = bw_entropy(),
  estimand = "ate"
)
study$w <- as.numeric(weights(fit))

outcome_mod <- glm(
  event ~ exposure,
  data = study,
  family = quasibinomial(),
  weights = w
)
```

The outcome model reports a standard error for the exposure coefficient, but it is conditional on the weights. It treats the reweighting as given and so leaves out one of the sources of variation in the estimate.

```{r naive}
naive <- summary(outcome_mod)$coefficients["exposure", ]
naive
```

The right standard error propagates the uncertainty from estimating the weights into the effect estimate. Whether it is larger or smaller than the naive value depends on the estimand and the design, but it is generally not the same, and using the naive value can give intervals with the wrong coverage.

## M-estimation with `ipw()`

The estimating-equation methods (`bw_entropy()`, `bw_ipt()`, and the just-identified `bw_cbps()`) fit weights that solve smooth estimating equations. That structure lets the uncertainty be propagated analytically, without resampling. `ipw()` writes one stacked estimating-equation system whose parameters are the weight parameters, the outcome-model coefficients, the two marginal means, and the effect contrasts, and the deli package differentiates that system at the fitted values and returns its empirical sandwich covariance. Nothing is re-solved: every parameter enters at the value its own fit already found. Because each contrast is a parameter of the stack rather than a transformation applied afterward, its standard error comes straight off the diagonal of the joint covariance, with no delta-method step in between. The result is a standard error that accounts for having estimated the weights.

```{r ipw}
result <- ipw(fit, outcome_mod)
result
```

The `rd` row is the risk difference on the same scale as the outcome model's marginal effect. Comparing its standard error to the naive one shows the correction.

```{r compare}
rd <- as.data.frame(result$estimates)
rd <- rd[rd$effect == "rd", ]

data.frame(
  source = c("naive (weights fixed)", "ipw (weights estimated)"),
  std.err = c(naive[["Std. Error"]], rd$std.err)
)
```

`ipw()` computes this variance for a binary exposure, an estimating-equation fit at exact balance, and a marginal outcome model whose only predictor is the exposure, the form the stacked system is derived for. The outcome model may be a `glm()` with a binomial, quasibinomial, or gaussian family, or a plain `lm()`. Any link those families carry is handled exactly, including a non-canonical one such as probit, because the sandwich differentiates the score it is given rather than assuming an information matrix. Model offsets are supported natively, written either as an `offset()` term in the formula or passed through the model's `offset` argument, and are carried through both the outcome-model score and the marginal means.

Everything else takes the bootstrap. The quadratic-program methods (`bw_energy()`, `bw_cfd()`, and `bw_sbw()`) and an entropy fit at a positive tolerance solve inequality-constrained problems, whose solutions are characterized by which constraints bind at the optimum rather than by a smooth system set to zero. An over-identified `bw_cbps()` fit minimizes a generalized-method-of-moments criterion in more moment conditions than it has parameters, so its balancing conditions are not solved to zero either. None of these has an estimating-equation representation to stack, so the fit carries none and `ipw()` raises an error pointing here. Inverse probability tilting is the exception to the tolerance rule: it solves the same smooth estimating equations whatever tolerance is requested, so `ipw()` handles it at any tolerance.

```{r unsupported, error = TRUE}
sbw_fit <- balance(
  study,
  exposure,
  c(age, score),
  method = bw_sbw(),
  constraints = balance_terms(tolerance = 0.02)
)
ipw(sbw_fit, outcome_mod)
```

## Bootstrapping the other fits

Where a fit carries no estimating equations, inference uses resampling instead. The recipe is a standard nonparametric bootstrap: resample the rows with replacement, refit the weights and the outcome model on each resample, and take the spread of the effect estimates across resamples as the standard error. The important detail is that the reweighting is refit inside the loop, so every source of variation, including the estimation of the weights, is captured. The example below uses `bw_sbw()`, but the recipe is the same for any fit and it also covers the cases `ipw()` declines for reasons other than the weights, such as a covariate-adjusted outcome model.

We keep the example small so it runs quickly: a few hundred resamples on a modest sample.

```{r bootstrap}
boot_study <- study[seq_len(400), ]

fit_effect <- function(data) {
  fit <- balance(
    data,
    exposure,
    c(age, score),
    method = bw_sbw(),
    constraints = balance_terms(tolerance = 0.02)
  )
  data$w <- as.numeric(weights(fit))
  model <- glm(
    event ~ exposure,
    data = data,
    family = quasibinomial(),
    weights = w
  )
  # The marginal risk difference by g-computation
  p1 <- predict(model, transform(data, exposure = 1), type = "response")
  p0 <- predict(model, transform(data, exposure = 0), type = "response")
  mean(p1) - mean(p0)
}

set.seed(1)
point_estimate <- fit_effect(boot_study)

n_boot <- 200
boot_estimates <- vapply(seq_len(n_boot), function(b) {
  rows <- sample(nrow(boot_study), replace = TRUE)
  fit_effect(boot_study[rows, ])
}, numeric(1))
```

The bootstrap standard error is the standard deviation of the resampled estimates, and a percentile interval comes from their quantiles.

```{r bootstrap-summary}
data.frame(
  estimate = point_estimate,
  std.err = sd(boot_estimates),
  ci.lower = quantile(boot_estimates, 0.025),
  ci.upper = quantile(boot_estimates, 0.975),
  row.names = NULL
)
```

A production analysis would use more resamples, typically at least one or two thousand, and might use the bias-corrected and accelerated interval rather than the percentile one. The structure stays the same: refit the weights inside every resample.

## Choosing an approach

Use `ipw()` when the method carries estimating equations, which is the estimating-equation family with a binary exposure and exact balance. It is faster and needs no tuning. Use the bootstrap for the quadratic-program methods, for an entropy fit at a positive tolerance, for an over-identified `bw_cbps()` fit, for covariate-adjusted outcome models, and for exposure types beyond binary. The bootstrap is more general but more expensive, and its precision improves with the number of resamples.
